\documentclass[10pt,a4paper]{article}
\usepackage{sltc2016}
\usepackage{url}
\usepackage{framed}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}
,fontsize=\small}
% Add ',fontsize=\small' for more characters per line
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}

\begin{document}

\date{\today}
\title{Outsourcing morphology in Grammatical Framework: a case study for Hungarian}
\name{Inari Listenmaa}
\address{Department of Computer Science and Engineering \\ University of Gothenburg and Chalmers University of Technology \\ \tt inari@chalmers.se}


\abstract{We implement a miniature resource grammar in Grammatical Framework (GF) by using 
resources developed in the Apertium community: a finite-state morphological 
transducer and a disambiguation grammar. 
Our goals are twofold: to share resources within the rule-based community, as 
well as to prevent the GF grammar growing in size. Especially for languages with 
complex morphology, not having to store large inflection tables makes the grammar
smaller and faster. As for development effort, we hope that the external resources
would also save time in the grammar writing process. The next steps are to scale up
to a full resource grammar, and parametrise the grammar for different tagsets.}

\maketitleabstract


\section{Introduction}
\label{sec:introduction}
Grammatical Framework (GF, \cite{ranta2004jfp}) is a grammar formalism and a programming 
language for writing multilingual grammars.
A GF grammar consists of an \emph{abstract syntax}, which declares the
categories and constructions in the grammar, and a number of \emph{concrete
  syntaxes}, where the categories and constructions are implemented,
separately for each languages.
By allowing multiple concrete syntaxes for a single abstract syntax, GF is a natural 
choice for interlingual translation.

One of the most important contributions of GF is its Resource Grammar Library 
\cite{ranta2009lilt}, which contains 40 languages as of September 2016. All the languages 
share the same core abstract syntax, and each language can have an extra module for 
constructions that are particular to that language. Via the common core, we can 
translate basic syntactic structures between any pair of the 40 languages; any new 
language added to the library will be connected to all of the existing languages.

Typically, writing a morphological description is the first step in starting a resource grammar.
A new grammar writer can easily spend weeks or months in defining inflection classes 
and writing morphophonological transformation rules for their language. 
Such a description in GF may well be valuable on its own right: it may provide 
insights about the morphological complexity of a language
\cite{detrez_ranta2012eacl}, or even lend itself for language description.
However, creating a morphological description in GF is time-consuming, and the result 
is often less efficient than a finite-state description.
If a new RGL language already has morphological resources, using them would ideally
speed up both development and performance of the resource grammar.
%\footnote{Of course, we are not denying anyone their fun of writing a morphological description in GF.}


\section{Implementation}

We implemented a miniature version (44 functions) of the GF resource
grammar \cite{ranta2009lilt} for this experiment. 
At this stage, the grammar is not suited for more than toy
applications, but the effort of grammar writing seems promising.
The development of this initial grammar took just a
couple of hours; the author has no knowledge of Hungarian, but years
of experience with GF and Apertium.

In the following sections, we present briefly the traditional way of writing GF
grammars, followed by the new method.

\subsection{Traditional GF grammar}

A self-contained GF grammar stores inflection tables in each lexical
entry. Then, the syntactic functions select the appropriate forms from
the tables. We will illustrate the process with a grammar that has
four categories: NP, V, VP and S, and two syntactic functions: complementation
and predication. The \texttt{Compl} function combines a NP and a V
into a VP, and \texttt{Pred} combines a NP and a VP into an S.
The abstract syntax of this grammar is shown below.
 


\begin{Shaded}
\begin{Highlighting}[]
  \NormalTok{abstract }\DataTypeTok{Grammar} \FunctionTok{=} \NormalTok{\{}

  \NormalTok{cat }
    \DataTypeTok{S} \NormalTok{; }\DataTypeTok{NP} \NormalTok{; }\DataTypeTok{VP} \NormalTok{; }\DataTypeTok{V} \NormalTok{;}

  \NormalTok{fun}
    \DataTypeTok{Pred} \FunctionTok{:} \DataTypeTok{NP} \OtherTok{->} \DataTypeTok{VP} \OtherTok{->} \DataTypeTok{S} \NormalTok{;}
    \DataTypeTok{Compl} \FunctionTok{:} \DataTypeTok{V} \OtherTok{->} \DataTypeTok{NP} \OtherTok{->} \DataTypeTok{VP} \NormalTok{;}
  \NormalTok{\}}
\end{Highlighting}
\end{Shaded}

% \begin{verbatim}
%   abstract Grammar = {

%   cat 
%     S ; NP ; VP ; V ;

%   fun
%     Pred : NP -> VP -> S ;
%     Compl : V -> NP -> VP ;
%   }
% \end{verbatim}

Next, we write a concrete syntax for a language where verbs inflect
for agreement and nouns for case. 
We introduce the parameters \texttt{Case} and \texttt{Agr}, and create
the categories specific to this language.


\begin{Shaded}
\begin{Highlighting}[]
  \NormalTok{param}
    \DataTypeTok{Case} \FunctionTok{=} \DataTypeTok{Nom} \FunctionTok{|} \DataTypeTok{Acc} \FunctionTok{|} \DataTypeTok{Dat} \FunctionTok{|} \FunctionTok{...} \NormalTok{;}
    \DataTypeTok{Agr} \FunctionTok{=} \DataTypeTok{SgP1} \FunctionTok{|} \DataTypeTok{SgP2} \FunctionTok{|} \FunctionTok{...} \FunctionTok{|} \DataTypeTok{PlP3} \NormalTok{;}
  \NormalTok{lincat }
    \DataTypeTok{S}  \FunctionTok{=} \DataTypeTok{Str} \NormalTok{;}
    \DataTypeTok{VP} \FunctionTok{=} \DataTypeTok{Agr} \OtherTok{=>} \DataTypeTok{Str} \NormalTok{;}
    \DataTypeTok{V} \FunctionTok{=} \NormalTok{\{ s }\FunctionTok{:} \DataTypeTok{Agr} \OtherTok{=>} \DataTypeTok{Str} \NormalTok{; compl }\FunctionTok{:} \DataTypeTok{Case} \NormalTok{\} ;}
    \DataTypeTok{NP} \FunctionTok{=} \NormalTok{\{ s }\FunctionTok{:} \DataTypeTok{Case} \OtherTok{=>} \DataTypeTok{Str} \NormalTok{; agr }\FunctionTok{:} \DataTypeTok{Agr} \NormalTok{\} ;}
\end{Highlighting}
\end{Shaded}

%\begin{verbatim}
%  param
%    Case = Nom | Acc | Dat | ... ;
%    Agr = SgP1 | SgP2 | ... | PlP3 ;
%  lincat 
%    S  = Str ;
%    VP = Agr => Str ;
%    V = { s : Agr => Str ; compl : Case } ;
%    NP = { s : Case => Str ; agr : Agr } ;
%\end{verbatim}

Then we implement the functions which operate on the given categories.
The complementation function must choose a right case from the object
NP, depending on the verb. Likewise, the predication function must
choose the right agreement from the VP, depending on the subject NP.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lin }
  \DataTypeTok{Compl} \NormalTok{v obj }\FunctionTok{=}
   \NormalTok{table \{ agr }\OtherTok{=>} \NormalTok{verb}\FunctionTok{.}\NormalTok{s }\FunctionTok{!} \NormalTok{agr}
               \FunctionTok{++} \NormalTok{obj}\FunctionTok{.}\NormalTok{s }\FunctionTok{!} \NormalTok{v}\FunctionTok{.}\NormalTok{compl \} ;}

  \DataTypeTok{Pred} \NormalTok{subj vp }\FunctionTok{=} \NormalTok{subj }\FunctionTok{!} \DataTypeTok{Nom} 
              \FunctionTok{++} \NormalTok{vp }\FunctionTok{!} \NormalTok{subj}\FunctionTok{.}\NormalTok{agr ;}
\end{Highlighting}
\end{Shaded}

% \begin{verbatim}
% lin 
%   Compl verb obj =
%    table { agr => verb.s ! agr
%                ++ obj.s ! verb.compl } ;

%   Pred subj vp = subj ! Nom 
%               ++ vp ! subj.agr ;

% \end{verbatim}

Because of these parameters, the grammar will produce only grammatical
sentences. Given the NP \emph{a fa} `the tree' and the verb
\emph{szeret} `love', the grammar will output \emph{a fa szereti a
  f\'{a}t} `the tree loves the tree'.

On the high level, this grammar design looks simple. 
But under the hood, the GF source code is compiled into a low-level
format. Adding a new value to a parameter or a new field to a record 
does not change much in the high-level rules, but each addition
multiplies the number of low-level rules. As a result, GF grammars for
morphologically complex languages often suffer from performance issues.

% Nouns, verbs and other inflecting categories contain large tables, for all
%  combinations of number and case. The verb, likewise, contains all the
%  conjugation, including tense, mood, person, aspect, and even
%  non-finite forms (e.g. {\em singing}). 

\subsection{GF grammar with external resources}

In the version with an external morphological analyser, all lexical
entries contain only a base form, along with a POS tag. The features,
which were used to select right forms from the inflection tables, are
now replaced with string fields for \emph{tags}, such as following.

\begin{Shaded}
\begin{Highlighting}[]
  \NormalTok{lincat}
    \DataTypeTok{S}  \FunctionTok{=} \DataTypeTok{Str} \NormalTok{;}
    \DataTypeTok{VP} \FunctionTok{=} \NormalTok{\{ verb }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{; obj }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{\} ;}
    \DataTypeTok{NP} \FunctionTok{=} \NormalTok{\{ s }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{; agrTag }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{\} ;}
    \DataTypeTok{V2} \FunctionTok{=} \NormalTok{\{ s }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{; complTag }\FunctionTok{:} \DataTypeTok{Str} \NormalTok{\} ;}
\end{Highlighting}
\end{Shaded}


Then, the syntactic functions can be reduced into concatenating tags,
and taking care of the right word order. Below are the same syntactic
functions for complementation and predication.


\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lin }
  \DataTypeTok{Compl} \NormalTok{v o }\FunctionTok{=} \NormalTok{\{ verb }\FunctionTok{=} \NormalTok{v}\FunctionTok{.}\NormalTok{s ;}
                \NormalTok{obj }\FunctionTok{=} \NormalTok{o}\FunctionTok{.}\NormalTok{s }\FunctionTok{+} \NormalTok{v}\FunctionTok{.}\NormalTok{complTag \} ;}

  \DataTypeTok{Pred} \NormalTok{subj vp }\FunctionTok{=} \NormalTok{(subj }\FunctionTok{+} \StringTok{"<nom>"}\NormalTok{)}
              \FunctionTok{++} \NormalTok{(vp}\FunctionTok{.}\NormalTok{verb }\FunctionTok{+} \NormalTok{subj}\FunctionTok{.}\NormalTok{agrTag)}
              \FunctionTok{++} \NormalTok{vp}\FunctionTok{.}\NormalTok{obj ;}
\end{Highlighting}
\end{Shaded}

% The entries in the GF grammar consist of just base lemmas with tags
% added: for example, the accusative and dative forms of the second
% person singular pronoun are stored as \texttt{te<prn><p2><sg><acc>}
% and \texttt{te<prn><p2><sg><dat>} instead of their inflected forms,
% \emph{téged} and \emph{neked} respectively.
With this setup, the GF trees are linearised into what looks like
Apertium analyses: for example,
\texttt{mi<prn><pers><p1><mf><pl><nom>} \texttt{ j\'{a}r<vblex><past><p3><pl>} for `we walked'.

\subsection{Pipeline}

The user of the grammar types in normal Hungarian words, and the input is analysed 
by the external morphological analyser, which is further disambiguated by a 
Constraint Grammar (CG). Only then is the sentence given to the GF grammar, 
which will return the syntactic parse tree.
The grammar can also beused in translation from another language
to Hungarian. First the source language is analysed into a GF tree,
then the tree is linearised into Hungarian with tags. To get actual
Hungarian, the external morphological analyser is called to generate
the inflected forms from the tags.



\section{Discussion}


Adding external tools into a GF grammar introduces more dependencies,
which makes the system more complex. In comparison, a 100 \% GF
grammar is easier to understand, and immune to the eventual changes in
the other systems. It is understandable if a grammar writer decides to
value independence and clarity over the convenience of reusing resources.
Furthermore, we have only implemented a fragment of the resource grammar:
scaling up to the full grammar may reveal challenges in the approach,
which would increase the development effort.

On a more speculative note, the GF grammar could work as a unification effort.
One of the most unnecessary annoyances in computational linguistics is
tagset incompatibility. To demonstrate, table~\ref{table:treeDoesntLove} shows two parses of the Hungarian sentence \emph{a fa nem szeret engem} 'the tree doesn't love me', in Ocamorph and Apertium.


\begin{table}[t!]
\center
    \begin{tabular}{l|l}
    Ocamorph tagset & Apertium tagset \\ \hline
    \texttt{a/ART}       & \texttt{a<det><def>}       \\
    \texttt{fa/NOUN}              & \texttt{fa<n><sg><nom>}            \\
    \texttt{nem/ADV}     &  \texttt{nem<adv>} \\
    \texttt{nem/UTT-INT} &  \texttt{nem<ij>}  \\
    \texttt{szeret/VERB}          & \texttt{szeret<vblex><pri><p3><sg>} \\
    \texttt{\'{e}n/NOUN<PERS<1>><CAS<ACC>>} & \texttt{\'{e}n<prn><pers><p1><mf><sg><acc>}  \\
    \end{tabular}
\label{table:treeDoesntLove}
\caption{Morphological analyses of the Hungarian sentence \emph{a fa
    nem szeret engem} `the tree doesn't love me'}
\end{table}

[TODO: functors for tagset variation]

\section{Conclusion}

Small scale experiment, very little effort so far. 
Should test correctness.
Nice showcase in sharing resources.


\bibliographystyle{sltc2016}
\bibliography{GF-RG.bib}


\end{document}
